Characterising cheque dishonour cases in India: Causes for delays and policy implications

### 2022-04-03 DOC/paper_chequeDishonour_combined.tex
Updating the paper.

\graphicspath{{../DOC/SLIDES/IMAGES/}}
%\graphicspath{{~/SLIDES/IMAGES/}}

@@

\usepackage[acronym,nonumberlist,nomain]{glossaries}
\makeglossary
%\loadglsentries{acronyms.tex}
\newglossaryentry{adr}{name={ADR},description={Alternative Dispute Resolution}}
\newglossaryentry{crpc}{name={CrPC},description={Code of Criminal Procedure}}
%\newacronym{adr}{ADR}{Alternative Dispute Resolution}
%\newacronym{crpc}{CrPC}{Code of Criminal Procedure}
\newacronym{coe}{CoE}{Committee of Experts}
\newacronym{lci}{LCI}{Law Commission of India}
\newacronym{ni}{NI Act}{Negotiable Instruments Act, 1881}
\newacronym{njdg}{NJDG}{National Judicial Data Grid}
\newacronym{sci}{SCI}{Supreme Court of India}
\newacronym{ut}{UT}{Union Territory}

@@

		This paper analyses the determinants of case disposition time under the Negotiable Instruments Act, 1881 by creating a unique dataset of roughly 50,000 case-level records spread across ten major states in India. Unlike previous studies, our analysis focuses mainly on case characteristics directly linked with the procedural matters to estimate the case duration. Regression results show that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, other jurisdictional issues, and reference to mediation significantly increases the case duration. Our findings are robust even when we consider the number of hearings as proxy for case disposition time. Additionally, results indicate that on average, cases being contested have a lower duration though it takes roughly three additional hearings to resolve. Our estimates also indicate that the 2015 amendment focused on clarifying the jurisdiction issues and mandatory centralization of cases against the same drawer has reduced the case duration by six months within a span of two years.

		%Our results also indicate that 2015 amendments focused on clarifying the jurisdiction related issues for filing cases and mandatory centralisation of cases against the same drawer has led to a major decrease of case duration. Our estimate suggest that it has reduced the case duration on average by more than six months within a span of two years.

		%Regression results show that cases referred for mediation on average takes 108 additional days to resolve. Additionally, results indicate that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, other jurisdictional issues

		%The findings are robust when we consider number of hearings as a proxy for case disposition time. We find 3.27 additional hearing for cases sent for mediation.

		%Furthermore, if we consider number of hearings as a proxy for case disposition time, we find 3.27 additional hearing for cases sent for mediation.

@@

	%\newpage
	%\printglossaries

	%\newpage
	%\listoftables

@@

	India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has increasingly become a top priority for Indian judges and policymakers.

	%India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. Slow judiciaries have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, minimising unnecessary judicial delays could help improve enforcement and enhance the overall rule of law. It is estimated that judicial delays cost India around 1.5\% of its GDP annually.\footcite{dey2016_cost} Against this backdrop, it is hardly surprising that tackling judicial delay has increasingly become a top priority for Indian judges and policymakers.

	%SOMETHING TO BE ADDED LATER: Link number of cases to state GDP. No of cases is directly linked to the economic activity. If delay hinders performance, then it will hurt the well performing states more. state GDP will suffer and it will have asymmetrical impact on Indian GDP. Insert the figure of SGDP and no of NI 138 cases.

@@

	Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \cite{sridhar2017_cheque} examined the behavior of cheque dishonor cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, she found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case-level data for eight states and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of cases that stay unresolved within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different states. As per our own estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most states.

%	The probability of case completion within 180 days is less than 5\%.
%	The probability of case completion within 270 days is 22\%.

	\begin{figure}[!ht]
		\centering
		\caption{Survival probability of cases across states}\label{fig:stateSurvival}
		\footnotesize
		\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days. 
%		\includegraphics[width = 0.9\textwidth]{/SLIDES/IMAGES/surv_states.pdf}
		\includegraphics[scale=0.35]{survstates.pdf}
	\end{figure}

	Since cheque dishonor cases are a large burden on the judiciary, various committees, institutions, and government bodies have thus focused on expediting disposal. For example, the \cite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonor cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \cite{bhan2015_placing}). Multiple amendments were also done in 2015 and 2018 to deal with the various procedural matters. The Supreme Court has also issued directions to the lower courts as well as various financial institutions to take a more pragmatic and realistic approach for speedy disposal of \gls{ni} cases.

	%However, it did not define \textit{fast-track courts} or give guidance concerning how and where they would operate.

	%Such knowledge is based on the assumption that cheque dishonour cases are a large burden on the judiciary and has thus focused on expediting disposal. For example, \gls{lci} (2008), by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts \autocite{lci2008_138, bhan2015_placing}. However, it did not define \textit{fast-track courts} or give guidance concerning how and where they would operate. Developing this work, \cite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases by looking at data from the DAKSH database across India’s courts for over 67,000 cases. In most cases, it found that resolution is delayed well beyond statutorily prescribed timelines and that certain banks and financial institutions are frequent complainants \autocite{sridhar2017_cheque}. As per our own estimates, less than 20\% of the cases get disposed within the stimulated timeline (see Figure \ref{fig:stateSurvival} for comparison across states in India).

@@

	%As \cref{sec:history} shows, most knowledge and development concerning such cases in India has come from government institutions, i.e. the \gls{lci} and the judiciary.

%	In 2020, the Supreme Court of India also took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court formed a \gls{coe}\footnote{Headed by Hon’ble Mr Justice RC Chavan, former Judge of the Bombay High Court.} and appointed Amici Curiae to assist the court,\footnote{Mr Sidharth Luthra (Sr Advocate) and Mr K Parameshwar (Advocate).} to study processes to expedite disposal of complaints under \S~138 of the \gls{ni}. This is in a series of interventions targeted to understand cheque dishonour cases in India. As \cref{sec:history} shows, most knowledge and development concerning such cases in India has come from government institutions, i.e. the \gls{lci} and the judiciary. The Amici appointed by the Supreme Court also made recommendations to expedite the disposal of cases. This included (i) increasing the use of pre- and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc.\footcite[For details see ]{amicus2020_submission}

	%\textbf{What we do?}:

	Given this background, the purpose of this paper is fourfold. First, using the case-level eCourts data, for the first time in India, we estimate the real judicial caseload due to the Negotiable Instruments Act, 1881 cases. Second, we estimate the average case disposition time and its variation over time and across states. 
Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposition time using regression analysis. In line with \cite{bielen2015}, using the interim and final orders data at the case level, we establish the link between the specific case characteristics and procedural issues, while controlling for characteristics that are inherent to courts and judges. Note that the empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case-level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targetted using procedural reforms.

%Given this background, the purpose of this paper is fourfold. First, using the case-level eCourts data, for the first time in India, we estimate the real judicial caseload due to the Negotiable Instruments Act, 1881 cases. Second, we estimate the average case disposition time and its variation over time and across states.  Third, we empirically test our conceptual framework of procedural determinants of case disposition time. In line with \cite{bielen2015}, using the interim and final orders data at the case level, we establish the link between the specific case characteristics and empirically examine the procedural issues and micro-level determinants of case duration, while controlling for characteristics that are inherent to courts and judges. Note that the empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case-level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targetted using procedural reforms.

%	Given access to the interim and final orders data, we focus on the reasons related to the procedural matters together with specific case characteristics that are relatively scarce, presumably as a consequence of lack of access to case-level data.

%And finally, we analyze the effect of the 2015 amendment of the NI Act, 1881 by comparing the duration of cases handled before and after their implementation using regression analysis.

%We also determine the number of hearing as an additional proxy for the trial length.

Additionally, we examine the role of the Negotiable Instruments (Amendment) Bill, 2015 which was passed by the Parliament in December 2015. The amendment was focused on clarifying the jurisdiction-related issues for filing cases for an offence committed under section 138 of the Negotiable Instruments Act, 1881. Also, it mandated the centralization of cases against the same drawer. These reforms were specially drafted to accelerate court proceedings even though no evidence-based insight into the causes of prolonged trials was available. We analyze the effect of the 2015 amendment of the NI Act, 1881 by comparing the duration of cases handled before and after their implementation using regression analysis. Our research makes it possible to evaluate the effectiveness of these reforms, at least for the states for which we gathered the data.

%Given this background, the purpose of this paper is fourfold. First, using the case-level eCourts data, for the first time in India, we estimate the real judicial caseload due to the Negotiable Instruments Act, 1881 cases. Second, we estimate the average case disposition time and its variation over time and across states. Third, we empirically test our conceptual framework to determine case-specific and procedural determinants of case disposition time. And finally, we analyze the effect of the 2015 amendment of the NI Act, 1881 by comparing the duration of cases handled before and after their implementation using regression analysis.

%Establishing insight into the case disposition time is indispensable for economic and judicial reform. Therefore, we first determine the duration of the cases included in our sample. We also determine the number of hearing as an additional proxy for the trial length. Subsequently, in line with \cite{bielenetal2015}, using the interim and final orders data at the case level, we empirically examine the procedural issues and micro-level determinants of case duration, while controlling for characteristics that are inherent to courts and judges. The results arising from such analyses will enable policymakers to set up cost-effective delay-reducing measures. Note that the empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case-level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the reasons related to the procedural matters together with specific case characteristics that are relatively scarce, presumably as a consequence of lack of access to case-level data.

%Additionally, we examine the role of the Negotiable Instruments (Amendment) Bill, 2015 which was passed by the Parliament in December 2015. The amendment was focused on clarifying the jurisdiction-related issues for filing cases for an offence committed under section 138 of the Negotiable Instruments Act, 1881. Also, it mandated the centralization of cases against the same drawer. These reforms were specially drafted to accelerate court proceedings even though no evidence-based insight into the causes of prolonged trials was available. Our research makes it possible to evaluate the effectiveness of these reforms, at least for the court where the data were gathered.

%This study, based on orders of the Supreme Court and the report of the Amici Curiae, attempts to assess the effect of suggested interventions on the expeditious trial of cases under \S~138 of the \gls{ni}. First, we empirically estimate the volume of NI 138 cases in differing courts. Given the disagreement, any acceptance among policymakers that \gls{ni} cases (\textit{cheque dishonor cases} hereinafter) are the reason for India's slow judiciary is likely to lead to misguided solutions. The misrepresentation could lead to diverting resources from other routes of judicial reform. Thus, one requires a more detailed analysis of the proportion of such cases, their causes, timelines, etc. This information can help judges and policymakers better target interventions.\footcite[For the importance of accurate judicial data, see][]{damle2020_ecourtsData, daksh2020_data, damle2020_land}

%\textbf{Key findings/Summary of our results}

Our estimates suggest that cheque dishonor cases represent approximately 13.2\% of courts' workload (pending and disposed) in India. This number could be an underestimate because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. 
%Some may be NI Act cases, even though the Act Name field does not say so.
We also found significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka (roughly in line with the court's recommendation) to as high as 547 days in Himachal Pradesh with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.

Controlling for state and year fixed effects, our findings suggest that cases where the accused fails to appear take an additional 213 days and 7 hearings to dispose of, all else being equal. Similarly, cases run as summons trials take 111 days and 7.2 hearings more than summary trials. Cases with jurisdictional issues take 287 days and 5.7 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 171 days and 10 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that 2015 amendments focused on clarifying the jurisdiction-related issues for filing cases and mandatory centralization of cases against the same drawer has led to significant reduction in case duration. Our estimates suggest that it has reduced the case duration on average by more than six months within a span of two years.

%	\begin{enumerate}
	%		\item Our estimates suggest that cheque dishonor cases represent approximately 13\% of courts' workload (pending and disposed) in India. This number could be an underestimate because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. Some may be NI Act cases, even though the Act Name field does not say so.
	%		\item We find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka (roughly in line with the court's recommendation) to as high as 547 days in Himachal Pradesh with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.
	%		\item Controlling for state and year fixed effects, our findings suggest that cases where the accused fails to appear to take an additional 213 days and 7 hearings to dispose of, all else being equal. Similarly, cases run as summons trials take 111 days and 7.2 hearings more than summary trials. Cases with jurisdictional issues take 287 days and 5.7 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 171 days and 10 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings.
	%		\item Our results also indicate that 2015 amendments focused on clarifying the jurisdiction-related issues for filing cases and mandatory centralization of cases against the same drawer has led to a major decrease in case duration. Our estimate suggests that it has reduced the case duration on average by more than six months within a span of two years.
%	\end{enumerate}
%			
	% SOMETHING TO BE ADDED LATER: \textbf{How we contribute to the debate?}

	The rest of this paper is organized as follows: after this introduction, \cref{sec:history} elaborates on the history of cheque dishonor provisions in India and their relation with judicial delays. In \cref{sec:select-case-char}, based on our conversations with practicing lawyers, other stakeholders, and the recommendations of the Amici Curiae and the Supreme Court, we figure out the key case characteristics that are linked to the procedural delay and can be directly targeted by the court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. In \cref{sec:2015amend}, we analyze the effectiveness of 2015 amendments by comparing the duration of cases handled before and after the procedural reforms were implemented. Finally, \cref{sec:conclusion} concludes the paper and presents the way forward.

	%	The rest of this paper is organized as follows: after this introduction, \cref{sec: history} elaborates on the history of cheque dishonor provisions in India and their relation with judicial delays. In \cref{sec:select-case-char}, based on our conversations with practicing lawyers, other stakeholders, and the recommendations of the Amici Curiae and the Supreme Court, we figure out the key procedural reforms and case-specific characteristics that can be targeted by direct court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. Finally, \cref{sec: conclusion} concludes the paper and presents the way forward.

@@

	The \acrlong{ni} was enacted to define the law relating to promissory notes, bills of exchange, and cheques. To increase the culture regarding the use of cheques and enhance the credibility of the instrument, the Act was amended in 1988.\footcite{niAmend1988} A new chapter (\S\S~138 to 142) was incorporated for penalties in the event of dishonor of cheques due to insufficient funds in the account of the drawer of the cheque. It also aimed to include adequate safeguards to prevent harassment of honest drawers. \S~138 provided for the circumstances under which a drawer can be penalized for the dishonor of a cheque (for a detailed discussion of \S~138, refer to Appendix \ref{app:understanding}).\footcite{ind1881_niAct}

	% SOMETHING TO BE DONE LATER: If the 1988 reform made cheque bouncing as a criminal offence, then this must be mentioned in the previous paragraph. Some literature must be cited on people's view on cheque bouncing cases being classified as a criminal offence.

	However, by 2001, these provisions were thought not to have had the desired effect.\footcite{stdcomm2001_138niAct} Large number of cases were pending across the country and the courts were not able to dispose of cases in a time-bound manner. The punishment was also thought to be inadequate. Given the large burden on courts, a Working Group was constituted the same year to review \S~138 of the \gls{ni} and make recommendations regarding the changes needed to effectively achieve the purpose of the section.\footcite{wg2001_138} In light of the recommendations of the Working Group, the government decided to bring further amendments to the Act. Among others, the amendments included:

@@

	The amendments aimed at speedy disposal of cheque dishonor cases through their summary trial and making the cases compoundable. The punishment provided under \S~138 also was enhanced from one year to two years. These legislative
	reforms aimed to encourage the usage of cheques so that regular business transactions and settlement of liabilities could be ensured. The amendments were considered by the Lok Sabha Standing Committee on Finance, which recommended that given the large burden on courts, the proposed amendments be coupled with the creation of specialized courts for \S~138 cases.\footcite{stdcomm2001_138niAct} However, the subsequent Amending Act did not reference specialised courts.\footcite{niAmend2002}

	The \gls{lci} took up the proposal for specialized courts in 2008. According to the Commission, the credibility of the financial sector was facing setbacks due to the large pendency of dishonored cheque cases. Relying on an array of judgments by the Supreme Court of India concerning speedy trials,\footcite{sc1978_khatoon, sc1981_champalal, sc2005_surinder, sc2008_krishna} the Commission recommended the introduction of fast track courts to address cases concerning dishonor of cheques under \S~138 of the Act. However, it did not comment on the number or expected workload of such courts.\footcite{lci2008_138} The need for additional courts was re-iterated by the \gls{lci} in 2009.\footcite{lci2009_reforms}

	While the \gls{lci} has focused on the need for more courts, the Supreme Court has assisted in implementing the amendments. In 2014, noting that the prime reason for the delays in \S~138 case is the absence of the accused, the Court held that the magistrate should adopt a pragmatic and realistic approach such as issuing notices and summons via e-mail or push notification to ensure delivery.\footcite{sc2014_iba} In 2018, the court directed banks to give the details of the e-mail of the accused to the complainant for service through e-mail. The court also directed that: (i) cases must be dealt with summarily, (ii) the evidence of the complainant must be conducted within three months, (iii) an endeavor must be made to conclude the trial within six months, (iv) the trial, as far as practicable, must be held on a day-to-day basis, and (v) High Courts may pass additional localized directions for speedy disposal of cases.\footcite{sc2018_meters}

	The Act was also amended in 2015 and 2018 to - clarify the appropriate area of the jurisdiction where cheque dishonor cases can be filed,\footcite{niAmend2015} and grant power to courts to grant interim compensation, respectively.\footcite{niAmend2018} In particular, the \citetitle{niAmend2015} 2015 aimed to address the difficulties faced by the payee or lender in filing cases under \S~138 due to a lack of clarity concerning the appropriate forum. It thus inserted specific provisions concerning the jurisdiction for an offence under \S~138, hoping to ensure that a fair trial of cases under \S~138 is conducted keeping in view the parties' interests by clarifying the territorial jurisdiction for trying cases.

	On the other hand, the \citetitle{niAmend2018} 2018 aimed to address the delays caused by unscrupulous drawers of dishonored cheques due to easy filing of appeals and obtaining a stay on proceedings. As a result, the payee would have to spend considerable time and resources in court proceedings to realize the value of the cheque. To discourage frivolous and unnecessary litigation, it inserted a new section to provide that a court trying an offence under section 138 may order the drawer of the cheque to pay interim compensation to the complainant, in a summary trial or a summons case, where he pleads not guilty to the accusation made in the complaint; and in any other case, upon framing of charge.

@@

	Noticeably, the recommendations of the Amici Curiae were in line with previously accepted reasons for delays in the disposal. While considering the recommendations, the Supreme Court recommended the Government and High Courts to take necessary action where possible and held that they should be the subject matter of deliberation by the \gls{coe} appointed in the same case.\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.}

@@

	The variables considered in the empirical literature on the length of court proceedings can be classified into three broad categories: (1) cross-country studies focused mainly on determinants at the country level (e.g., public budgets for courts, per capita GDP, etc.) (2) within-country variations in court performance focused on court characteristics (e.g. backlogs, number of judges, etc.) and (3) case-level studies focused on the case-specific characteristics like the number of parties involved, number of pleadings, etc. However, the procedural reforms target features that hinder the operational functioning of the court. These characteristics are difficult to gauge and empirical literature on these determinants remains scarce, presumably as a consequence of lack of access to case-level data. If we look at the past recommendations of the \gls{lci}, the Amici Curiae, the Supreme Court, and various other committees and reports, the target of these interventions fall into five broad categories:

	%The recommendations of the Amici Curiae and the Supreme Court target several aspects of a \gls{ni} case. They posit that these aspects likely lead to delays. This study examines six areas that are the targets of these interventions, viz. ---

>>>>

\graphicspath{{../DOC/SLIDES/IMAGES/}}

@@

\usepackage[acronym,nonumberlist,nomain]{glossaries}
\newglossaryentry{adr}{name={ADR},description={Alternative Dispute Resolution}}
\newglossaryentry{crpc}{name={CrPC},description={Code of Criminal Procedure}}
\newacronym{coe}{CoE}{Committee of Experts}
\newacronym{lci}{LCI}{Law Commission of India}
\newacronym{ni}{NI Act}{Negotiable Instruments Act, 1881}
\newacronym{njdg}{NJDG}{National Judicial Data Grid}
\newacronym{sci}{SCI}{Supreme Court of India}
\newacronym{ut}{UT}{Union Territory}

@@

This paper analyses the determinants of case disposition time under the Negotiable Instruments Act, 1881 by creating a unique dataset of roughly 50,000 case level records spread across ten major States in India. Unlike previous studies, this analysis focuses mainly on case characteristics directly linked with procedural matters to estimate the case duration. Regression results show that multiplicity of proceedings, non-appearance of accused, trials being run as summons trial, jurisdictional issues, and reference to mediation significantly increase the case duration. These findings are robust even when considering the number of hearings as a proxy for case disposition time. Additionally, results indicate that, on average, contested cases have a lower duration. However, it takes roughly three additional hearings to resolve. These estimates also indicate that the 2015 amendment to the Negotiable Instruments Act, 1881, which focused on clarifying the jurisdiction issues, and mandatory centralisation of cases against the same drawer, has reduced the case duration by six months.

@@

@@

India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has become a top priority for Indian judges and policymakers.

@@

Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \cite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, she found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case level data for eight States and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of unresolved cases within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different States. As per our estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most States.

\begin{figure}[!ht]
\centering
\caption{Survival probability of cases across States}\label{fig:stateSurvival}
\footnotesize
\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days.
\includegraphics[scale=0.35]{survstates.pdf}
\end{figure}

Since cheque dishonour cases are a large burden on the judiciary, various committees, institutions, and government bodies have focused on expediting disposal. For example, the \cite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \cite{bhan2015_placing}). The Act was also subject to amendments in 2015 and 2018 to deal with the various procedural matters. Similarly, the Supreme Court has issued directions to lower courts and various financial institutions to take a more pragmatic and realistic approach for the speedy disposal of \gls{ni} cases.

@@

Given this background, the purpose of this paper is fourfold. First, using case level eCourts data, for the first time in India, we estimate the real judicial caseload due to the \gls{ni}. Second, we estimate the average case disposal time and its variation over time and across States. Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposal time using regression analysis. In line with \cite{bielen2015}, using case level interim and final orders data, we establish the link between the specific case characteristics and procedural issues while controlling for inherent court and judge characteristics. The empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targeted using procedural reforms.

Additionally, we examine the role of the Negotiable Instruments (Amendment) Bill, 2015, which was passed by the Parliament in December 2015. The amendment focused on clarifying the jurisdiction-related issues for filing cases for an offence committed under \S~138 of the Negotiable Instruments Act, 1881. It also mandated the centralisation of cases against the same drawer. These reforms were specially drafted to accelerate court proceedings even though no evidence-based insight into the causes of prolonged trials was available. We analyse the effect of the 2015 amendment by comparing the duration of cases handled before and after their implementation using regression analysis. Our research makes it possible to evaluate the effectiveness of these reforms, at least for the States for which we gathered the data.

Our estimates suggest that cheque dishonour cases represent 13.2\% of courts' workload (pending and disposed) in India. This number could be underestimated because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. We also find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka to as high as 547 days in Himachal Pradesh, with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.

Controlling for State and year fixed effects, our findings suggest that, all else being equal, cases where the accused fails to appear take an additional 201 days and 7 hearings to dispose of. Similarly, cases run as summons trials take 112 days and 7.4 hearings more than summary trials. Cases with jurisdictional issues take 271 days and 5.5 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 168 days and 9.9 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that the 2015 amendments have significantly reduced case duration -- reducing the case duration by more than six months.

The rest of this paper is organised as follows: after this introduction, \cref{sec:history} elaborates on the history of cheque dishonour provisions in India and their relation to judicial delays. In \cref{sec:select-case-char}, based on our conversations with practising lawyers, other stakeholders, and the recommendations of the Amici Curiae and the Supreme Court, we figure out the key case characteristics linked to the procedural delay that can be directly targeted by court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. In \cref{sec:2015amend}, we analyse the effectiveness of the 2015 amendments by comparing the duration of cases handled before and after the procedural reforms were implemented. Finally, \cref{sec:conclusion} concludes the paper and presents the way forward.

@@

The \acrlong{ni} was enacted to define the law relating to promissory notes, bills of exchange, and cheques. To increase the culture regarding the use of cheques and enhance the credibility of the instrument, the Act was amended in 1988.\footcite{niAmend1988} A new chapter (\S\S~138 to 142) was incorporated for penalties in the event of dishonour of cheques due to insufficient funds in the account of the drawer of the cheque. It also aimed to include adequate safeguards to prevent harassment of honest drawers. \S~138 provided for the circumstances under which a drawer can be penalised for the dishonour of a cheque (for a detailed discussion of \S~138, refer to Appendix \ref{app:understanding}).\footcite{ind1881_niAct}

However, by 2001, these provisions were thought not to have had the desired effect.\footcite{stdcomm2001_138niAct} Large number of cases were pending across the country, and courts could not dispose of cases in a time-bound manner. The punishment was also thought to be inadequate. A Working Group was constituted the same year to review \S~138 of the \gls{ni} and make recommendations regarding the changes needed to effectively achieve the section's purpose.\footcite{wg2001_138} In light of the recommendations of the Working Group, the government decided to bring further amendments to the Act. Among others, the amendments included:

@@

The amendments aimed at speedy disposal of cheque dishonour cases through summary trial and making the cases compoundable. The punishment provided under \S~138 also was enhanced from one year to two years. These legislative
reforms aimed to encourage the usage of cheques so that regular business transactions and settlement of liabilities could be ensured. The amendments were considered by the Lok Sabha Standing Committee on Finance, which recommended that given the large burden on courts, the proposed amendments be coupled with the creation of specialised courts for \S~138 cases.\footcite{stdcomm2001_138niAct} However, the subsequent Amending Act did not reference specialised courts.\footcite{niAmend2002}

The \gls{lci} took up the proposal for specialised courts in 2008. According to the Commission, the credibility of the financial sector was facing setbacks due to the large pendency of cheque dishonour cases. Relying on an array of judgments by the Supreme Court of India concerning speedy trials,\footcite{sc1978_khatoon, sc1981_champalal, sc2005_surinder, sc2008_krishna} the Commission recommended the introduction of fast track courts to address cases concerning dishonour of cheques under \S~138 of the Act. However, it did not comment on such courts' number or expected workload.\footcite{lci2008_138} The need for additional courts was re-iterated by the \gls{lci} in 2009.\footcite{lci2009_reforms}

While the \gls{lci} has focused on the need for more courts, the Supreme Court has assisted in implementing the amendments. In 2014, noting that the prime reason for delays in \S~138 case was the absence of the accused, the Court held that the magistrate should adopt a pragmatic and realistic approach such as issuing notices and summons via e-mail or push notification to ensure delivery.\footcite{sc2014_iba} In 2018, the court directed banks to give the details of the e-mail of the accused to the complainant for service through e-mail. The court also directed that: (i) cases must be dealt with summarily, (ii) the evidence of the complainant must be conducted within three months, (iii) an endeavour must be made to conclude the trial within six months, (iv) the trial, as far as practicable, must be held on a day-to-day basis, and (v) High Courts may pass additional localised directions for speedy disposal of cases.\footcite{sc2018_meters}

The Act was also amended in 2015 and 2018 to - clarify the appropriate area of the jurisdiction where cheque dishonour cases can be filed,\footcite{niAmend2015} and grant power to courts to grant interim compensation, respectively.\footcite{niAmend2018} In particular, the \citetitle{niAmend2015} 2015 aimed to address the difficulties the payee or lender faced in filing cases under \S~138 due to a lack of clarity concerning the appropriate forum. It thus inserted specific provisions concerning the jurisdiction for an offence under \S~138, hoping to ensure that a fair trial of cases under \S~138 is conducted keeping in view the parties' interests by clarifying the territorial jurisdiction for trying cases.

On the other hand, the \citetitle{niAmend2018} 2018 aimed to address the delays caused by unscrupulous drawers of dishonoured cheques due to easy filing of appeals and obtaining a stay on proceedings. As a result, the payee would have to spend considerable time and resources in court proceedings to realise the value of the cheque. To discourage frivolous and unnecessary litigation, it inserted a new section to provide that a court trying an offence under \S~138 may order the drawer of the cheque to pay interim compensation to the complainant in a summary trial or a summons case, where he pleads not guilty to the accusation made in the complaint; and in any other case, upon framing of charge.

@@

Noticeably, the recommendations of the Amici Curiae were in line with previously accepted reasons for delays in the disposal. While considering the recommendations, the Supreme Court recommended the Government and High Courts take necessary action where possible and held that they should be the subject matter of deliberation by the \gls{coe} appointed in the same case.\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.}

@@

The variables considered in the empirical literature on the length of court proceedings can be classified into three broad categories: (1) cross-country studies focused mainly on determinants at the country level (e.g., public budgets for courts, per capita GDP, etc.) (2) within-country variations in court performance focused on court characteristics (e.g. backlogs, number of judges, etc.) and (3) case level studies focused on the case-specific characteristics like the number of parties involved, number of pleadings, etc. However, procedural reforms target features that hinder the operational functioning of the court. These characteristics are difficult to gauge, and empirical literature on these determinants remains scarce, presumably due to a lack of access to case level data. If we look at past recommendations by the \gls{lci}, the Amici Curiae, the Supreme Court, and various other committees and reports, the target of these interventions fall into five broad categories:

\begin{enumerate}
\item how to bring the accused to the court?;
\item how to run the trials (summary or summons trials)?;
\item whether to refer the case to mediation;
\item how to clarify jurisdictional issues?; and
\item how to avoid the multiplicity of proceedings?
\end{enumerate}
